---
title: Analysis of a common name in the U.S.
author: Josue Rodriguez
date: '2018-05-24'
slug: analysis-of-a-common-name-in-the-u-s
categories:
  - RMarkdown
  - ggplot2
tags: []
---

``` {r, echo=FALSE}
load('~/GitHub/Name-Analysis/name_analysis.RData')
```

# Introduction

This is just test post. Several months ago, I looked at how common a sex ambiguous name, "Taylor", is in the United States by year (1900 - 2000) and by sex.

# Results

## Prevalence by Year
Below are plots that show the proportion (log base 10 transformed) of people named "Taylor" over time. Popularity for the name "Taylor" began rising around the 60s before dipping back down in the 90s.

``` {r, fig.align='center', echo = FALSE}
year.plot

sex_year.plot
```

To test if there is a significant effect of year a linear model was built. Proportion of the population (log base 10 transformed) was the dependent variable and `year` was the independent variable. Results of this analysis indicated that `year` was a significant predictor for proportion of babys named "Taylor", and the coefficients show a positive slope.

``` {r}
summary(year.lm)
```

## Prevalence by Sex
The below plot shows the difference in proportion of babys named "Taylor" between males and females.

``` {r, fig.align='center', echo = FALSE}
sex_box.plot
```

Sex was not a significant predictor of babys being named "Taylor" between 1900 and 2000, although the coefficients also indicated a positive slope.

``` {r}
summary(sex.lm)
```