---
title: "Tutorial: Bayesian Testing of Central Structures"
author: "Josue Rodriguez"
date: "5/8/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/josue/Box/formalizing-expectations/04-tutorial-blog")

```
```{r include=FALSE}
source("01-plot-adj.R")
```


# Introduction

This tutorial demonstrates how to use networks (specifically Gaussian graphical models or GGMs) to 

1. Generate hypotheses
2. Perform confirmatory tests on the generated hypotheses

In psychology, network models are almost never used to generate hypotheses. This is puzzling because one of the original reasons researchers began using them was exactly this. Networks are also thought of highly exploratory tools, despite their potential for confirmatory tests. However, there is plenty of room for networks in this respect. These ideas are expanded upon in our recent paper,"On Formalizing Theoretical Expectations: Bayesian Testing of Central Structures in Pyschological Networks", where we merge exploratory and confirmatory hypotheses into a cohesive framework based on Bayesian hypothesis testing. You find a preprint [here](https://psyarxiv.com/zw7pf/).


In what follows, I will describe how you can use `R` to perform confirmatory hypothesis tests bases on an initial, exploratoy hypotheses with GGMs. For clarity, some code chunks have been omitted, but the full code to reproduce this document is available on the [OSF](https://osf.io/bh783) or [GitHub](https://github.com/josue-rodriguez/formalizing-expectations).

# Examples


## PTSD Network

To begin we need several packages:

* `BGGM`: to conduct exploratory and confirmatory analyses with GGMs
* `MASS`: to generate data from covariance matrices
* `networktools`: to calculate bridge centrality statistics



```{r}
# Uncomment and run  if missing packages

# for latest version of BGGM
# remotes::install_github('donaldrwilliams/BGGM') 

# packages <- c("BGGM", "MASS", "networktools")
# if (!packages %in% installed.packages()) install.packages(packages)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(MASS)
library(networktools)
library(BGGM)
```

```{r setup-ptsd, include=FALSE}
# column names
col_names <-
  c(
    # Re-experiencing
    "Intrusive Thoughts",
    "Nightmares",
    "Flashbacks",
    "Physiological/psychological reactivity",
    # Avoidance
    "Avoidance of thoughts",
    "Avoidance of situations",
    "Amnesia",
    "Disinterest in activities",
    "Feeling detached",
    "Emotional numbing",
    "Foreshortened future",
    # Arousal
    "Sleep problems",
    "Irritability",
    "Concentration problems",
    "Hypervigilance",
    "Startle response"
  )

# node names
node_names <- c(
  paste0("B", 1:4), # re-experienceing
  paste0("C", 1:7), # avoidance
  paste0("D", 1:5) # arousal
)
# community names
comms <- c(
  rep("Re-Experiencing", 4),
  rep("Avoidance", 7),
  rep("Arousal", 5)
)

# color palette
pal <- ggthemes::few_pal()(3)
node_cols <- c(
  "Re-Experiencing" = pal[1],
  "Avoidance" = pal[2],
  "Arousal" = pal[3])
```


The first dataset contains measurements on 16 PTSD symptoms in 3 communities, "Re-experiencing", "Avoidance", and "Arousal". Only the covariance matrices are available so we have to generate the data using `MASS`. The data,`ptsd_cor4`, is loaded when we load `BGGM`.

```{r gen-ptsd, include = FALSE}
set.seed(1812)
# data for exploratory analyses
explore_ptsd <- mvrnorm(n = 965, mu = rep(0, 16), Sigma = ptsd_cor4, empirical = TRUE)
colnames(explore_ptsd) <- node_names
```



### Exploratory analysis

#### Estimate graph

Our approach begins by estimating an exploratory network. With `BGGM`, this requires calling the `explore` function to obtain and sample the posterior distribution. These results are saved in `explore_network`. The `select` function takes the results from an `explore` call, and is used to determine the edge set for $\mathbf{A}^{CD}$ and $\mathbf{A}^{CI}$ --- the conditional (in)dependence structures.

Note that we formally incorporate a one-sided hypothesis test. $\mathcal{H}_0: \rho_{ij} = 0$ versus $\mathcal{H}_1: \rho_{ij} > 0$ by setting `alternative = "greater"` to determine the edge set. This formally includes the expectation of a "positive manifold", or that the edges should all be positive. 

```{r exp-ptsd}
# sample posterior distribution
explore_network <- explore(explore_ptsd, chains = 4, cores = 4)

# determine edge set
selected_network <- select(explore_network, alternative = "exhaustive", bf_cut = 3)
```


We can visualize the conditional (in)dependence structures by simply calling `plot`. By default, this returns three plots, but we focus on the first two.

```{r plot-ptsd-1, warning=FALSE, include=FALSE, eval=FALSE}
plots <- 
  plot(selected_network,
       node_labels = node_names, # vector storing node names
       communities = comms, # vector of community name for each node
       node_size = 12)

plots$plt_pos +
  ggtitle("Conditional Dependence Structure")

plots$plt_null +
  ggtitle("Conditional Independence Structure")
```



#### Bridge Centrality

We then proceed to calculate bridge strength using the `networktools` package. This is similar to node strength, in that, for a given node, it is the sum of the absolute values of its edges. However, bridge strength only takes into account edges that connect a node to different communities, or clusters. Thus, it is a measure of inter-community connectivity, and we use it here to identify central structures in a network. 

```{r bridge-ptsd}
# extract partial correlations from selected network
partial_cors <- selected_network$pcor_mat * selected_network$pos_mat

# rename columns with node names
dimnames(partial_cors) <- list(node_names, node_names)


# calculate bridge strength. comms is a vector specifying 
# the community for each node
bridge_strengths <- bridge(partial_cors, communities = comms)$`Bridge Strength`

# we use the top 10% in bridge strength as bridge nodes
bridge_strength_cutoff <- quantile(bridge_strengths, 0.9)

bridge_strengths[bridge_strengths > bridge_strength_cutoff]
```

Calculating bridge strength indicates that nodes B4 and D1 are the top bridge nodes.

#### Plot bridges


A key idea in our paper was that highlighting and "zooming" in on central structures allows researchers to easily formulate hypotheses. For example, we plotted the neighborhood of bridge relations for nodes D1 and B4 


```{r plot-ptsd-cd, include=FALSE, eval=FALSE}
plot_cd <- 
  make_adj_plot(selected_network$pos_mat, # conditional dependence structure
              node_labels = node_names, # vector storing node names
              communities = comms,
              bridges = c("D1", "B4"), # name of nodes to highlights
              fade = c("D2", "D5", "B3", # name of nodes to fade
                       "C3", "C4", "C5"),
              node_size = 12,
              type = "cd")  + # type = "cd" for conditional dependence. "ci" for conditional
  ggthemes::scale_fill_few() +
  theme_facet() +
  theme(legend.position = "top",
        legend.text = element_text(size = 14),
        legend.title = element_blank())

plot_cd
```

```{r plot-ptsd-ci, include=FALSE, eval=FALSE}
plot_ci <-
  make_adj_plot(selected_network$null_mat,
              node_labels = node_names,
              communities = comms,
              bridges = paste0("D", 1),
              fade = NULL,
              type = "ci", # type = "ci" for conditional independence
              node_size = 12) +
  ggthemes::scale_fill_few() +
  guides(fill = guide_legend(override.aes = list(size=10))) +
  theme_facet() +
  theme(legend.position = "top",
        legend.box.background = element_blank(),
        legend.background = element_blank(),
        legend.title = element_blank())
plot_ci
```


```{r plot-ptsd-bridges, echo=FALSE}
coord_d1 <- 
  data.frame(
    x = c(1, 0, 1, 2, 3, 4, 3),
    y = c(1, 2, 3, 2, 1, 2, 3),
    comm = c("3",  "3",  "3",  "1",  "2",  "2",  "2"),
    node = c("B1", "B4", "B2", "D1", "C6", "C1", "C2")
    )
    

plot_d1 <-
  ggplot(coord_d1, aes(x, y)) +
  # bridge cors
  annotate(geom = "segment",
           x = rep(2, 4),
           xend = c(1, 1, 3, 3),
           y = rep(2, 4),
           yend = c(1, 3, 1, 3),
           linetype = "dashed",
           size = 1) +
  # independent
  annotate(geom = "segment",
           x = c(2, 2),
           xend = c(0, 4),
           y = c(2, 2),
           yend = c(2, 2),
           linetype = "dotted",
           col = "gray70") +
  # B nodes
  annotate(geom = "segment",
           x = c(1, 1, 0),
           xend = c(0, 1, 1),
           y = c(1, 1, 2),
           yend = c(2, 3, 3),
           linetype = c(3, 1, 1),
           col = "gray70") +
  # C nodes
  annotate(geom = "segment",
           x = c(3, 3, 4),
           xend = c(4, 3, 3),
           y = c(1, 1, 2),
           yend = c(2, 3, 3),
           linetype = c(1, 3, 3),
           col = "gray70") +
  # bridge cor labels
  annotate(geom = "label",
           label = c(0.11, 0.31, 0.08, 0.15),
           x = c(1.5, 1.5, 2.5, 2.5),
           y = c(1.5, 2.5, 1.5, 2.5)) +
  geom_point(size = 21) +
  geom_point(aes(fill = comm), 
             size = 20, 
             shape = 21,
             show.legend = FALSE) +
  geom_text(aes(label = node)) +
  lims(x = c(-0.1, 4.1), y = c(0.9, 3.1))  +
  ggthemes::scale_fill_few() +
  theme_facet()


coord_b4 <- 
  data.frame(
    x = coord_d1$x,
    y = coord_d1$y,
    comm = c("2", "2", "2", "3", "1", "1", "1"),
    node = c("C7", "C6", "C1", "B4", "D4", "D1", "D3")
    )
    

plot_b4 <-
  ggplot(coord_b4, aes(x, y)) +
  # bridge cors
  annotate(geom = "segment",
           x = rep(2, 4),
           xend = c(1, 1, 3, 3),
           y = rep(2, 4),
           yend = c(1, 3, 1, 3),
           size = 1) +
  # independent
  annotate(geom = "segment",
           x = c(2, 2),
           xend = c(0, 4),
           y = c(2, 2),
           yend = c(2, 2),
           linetype = "dotted",
           col = "gray70") +
  # C nodes
  annotate(geom = "segment",
           x = c(1, 1, 0),
           xend = c(0, 1, 1),
           y = c(1, 1, 2),
           yend = c(2, 3, 3),
           linetype = c(3, 3, 1),
           col = "gray70") +
  # D nodes
  annotate(geom = "segment",
           x = c(3, 3, 4),
           xend = c(4, 3, 3),
           y = c(1, 1, 2),
           yend = c(2, 3, 3),
           linetype = c(1, 1, 3),
           col = "gray70") +
  # bridge cor labels
  annotate(geom = "label",
           label = c(0.11, 0.21, 0.07, 0.14),
           x = c(1.5, 1.5, 2.5, 2.5),
           y = c(1.5, 2.5, 1.5, 2.5)) +
  geom_point(size = 21) +
  geom_point(aes(fill = comm), 
             size = 20, 
             shape = 21,
             show.legend = FALSE) +
  geom_text(aes(label = node)) +
  lims(x = c(-0.1, 4.1), y = c(0.9, 3.1))  +
  ggthemes::scale_fill_few() +
  theme_facet()

plot_d1
plot_b4
```



### Confirmatory analysis

With the central structures identified and plotted, we can move on to formulating and testing hypotheses. 

Note that we generate another dataset here so that we test our hypotheses on different data than our exploratory analysis.

```{r conf-ptsd}
set.seed(1)

confirm_ptsd <- mvrnorm(n = 926, mu = rep(0, 16), Sigma = ptsd_cor3, empirical = TRUE)
colnames(confirm_ptsd) <- node_names
```


#### Varying degrees of replication
We first focus on node B4 and test the following hypotheses
\begin{align}
    \mathcal{H}_1&: (\rho_{B4-C1}, \rho_{B4-C7}, \rho_{B4-D3},  \rho_{B4-D4}) > 0 \\ \nonumber
    \mathcal{H}_2&: \rho_{B4-C1} > (\rho_{B4-C7}, \rho_{B4-D3}, \rho_{B4-D4}) > 0 \\ \nonumber
    \mathcal{H}_3 &: ``\text{not}\; \mathcal{H}_1 \; \text{or}\; \mathcal{H}_2  \text{.''}
\end{align}

Above, $\mathcal{H}_1$ is testing for replication of all edges but is otherwise agnostic towards the interplay among bridge relations. $\mathcal{H}_2$ then provides a refined view into the bridge neighborhood by testing an additional constraint that the strongest edge replicated. That is, all of the bridge relations \emph{and} the strongest edge re-emerged in an independent dataset. Furthermore, $\mathcal{H}_1$ and $\mathcal{H}_2$ both reflect a positive manifold. We also included $\mathcal{H}_3$ which accounts for structures that are not $\mathcal{H}_1$ or $\mathcal{H}_2$.


To test these hypotheses, we can formulate them in a single string and use the `confirm` function. Note that hypotheses are separated by a semicolon, and that partial correlations are denoted as `node1 -- node2`. The output is obtained by simply printing out the resuls of `confirm`.


```{r, hyp-var-rep, message=FALSE}
hyp_var_rep <- c("(B4--C1, B4--C7, B4--D3, B4--D4) > 0;
                   B4--C1 > (B4--C7, B4--D3, B4--D4) > 0")

confirm_var_rep <- confirm(confirm_ptsd,
                           hyp_var_rep,
                           iter = 50000)
confirm_var_rep
```

The output includes both the posterior probabilities and all the Bayes factors. The Bayes factors are in reference to the rows relative to the columns. For example the element in the 2nd row and 1st column would be interpreted as $\textbf{BF}_{21} = 4.05$

In this case, $\mathcal{H}_2$ is the preferred hypothesis, that is, all of the bridge edges and the strongest edge replicated.  This gets at an important notion. It is possible to test _varying degrees of replication_.

<!-- This analysis also indicates that (1) the bridge relations replicated in an independent dataset; and (2) the relation between ``sleep problems'' (node B4) and ``avoidance of thoughts'' (node C1) \emph{could} be the strongest bridge between the Re-experiencing and Avoidance communities. -->

## Comorbidity Network

We also examined a comorbity network containing 16 symptoms of anxiety and depression, and followed the same steps as above. This dataset is available on the [OSF](https://osf.io/wemcg/). Here, however, we split the data into two because we did not have independent datasets. We formulated hypotheses on one half and tested them on the remaining half. 

### Exploratory analysis

```{r setup-anxdep, include=FALSE}
# libraries
set.seed(27)

cov_anxdep <- read.csv("../05-data/00-cov-anxdep.csv")[, -1]

sim_anxdep <- MASS::mvrnorm(n = 1029, mu = rep(0, 16), Sigma = cov_anxdep, empirical = TRUE)

col_names <- 
  sub(pattern = "[A-Z].",
      replacement = "",
      x = colnames(cov_anxdep)) 

col_names <- tolower(col_names)

colnames(sim_anxdep) <- col_names
node_names <- 
  c(
    paste0("D", 1:9),
    paste0("A", 1:7)
  )

comms <- c(
  rep("depression", 9),
  rep("anxiety", 7)
)


pal <-  ggthemes::colorblind_pal()(8)[c(4, 6)]
split <- sample(1:1029, size = floor(1029 * .5))
explore_anxdep <- sim_anxdep[split, ]
confirm_anxdep <- sim_anxdep[-split, ]
colnames(confirm_anxdep) <-  node_names
```

```{r, explore-anxdep}
explore_network <- explore(explore_anxdep, chains = 4, cores = 4)
selected_network <- select(explore_network, alternative = "exhaustive", BF_cut = 3)
```

```{r explore-visual}
partial_cors <- selected_network$pcor_mat * selected_network$pos_mat

partial_cors <- round(partial_cors, 4)
dimnames(partial_cors) <- list(node_names, node_names)
```

#### Bridge Strength

```{r bridge, warning=FALSE}
bridge_strengths <- sort(bridge(partial_cors, communities = comms)$`Bridge Strength`, decreasing = TRUE) 
bridge_strength_cutoff <- quantile(bridge_strengths, 0.9)
bridge_strengths[bridge_strengths > bridge_strength_cutoff]
```

#### Plot bridges

```{r plot-anxdep, echo=FALSE}
coords <- 
  data.frame(
    x = c(1, 1, 2, 2, 3, 3),
    y = c(1, 4, 2, 3, 1, 4),
    comm = c("A", "A", "D", "D", "A", "A"),
    node = c("A5", "A3", "D8", "D6", "A7", "A6")
    )
    

plot_bridges <-
  ggplot(coords, aes(x, y)) +
  # bridge cors (d8 first)
  annotate(geom = "segment",
           x = rep(2, 4),
           xend = c(1, 3, 1, 3),
           y = c(2, 2, 3, 3),
           yend = c(1, 1, 4, 4),
           linetype = c(2, 2, 1, 1),
           size = 1) +
  # # independent
  annotate(geom = "segment",
           x =    c(2, 2, 2, 2, 2),
           xend = c(2, 1, 3, 1, 3),
           y =    c(2, 2, 2, 3, 3),
           yend = c(3, 4, 4, 1, 1),
           linetype = "dotted",
           col = "gray70") +
  # A nodes
  annotate(geom = "segment",
           x = c(1, 1, 3),
           xend = c(1, 3, 3),
           y = c(1, 4, 4),
           yend = c(4, 4, 1),
           linetype = c(3, 3, 1),
           col = "gray70") +
  # # bridge cor labels
  annotate(geom = "label",
           label = c("0.17", "0.10", "0.20", "0.20"),
           x =     c(1.5, 2.5, 1.5, 2.5),
           y =     c(3.5, 3.5, 1.5, 1.5)) +
  geom_point(size = 21) +
  geom_point(aes(fill = comm), 
             size = 20, 
             shape = 21,
             show.legend = FALSE) +
  geom_text(aes(label = node), col = "white") +
  lims(x = c(0.9, 3.1), y = c(0.8, 4.2))  +
  scale_fill_manual(values = pal) +
  theme_facet()

plot_bridges
```


### Confirmatory analysis



#### Intra- and Inter-Bridge Sets

The following hypotheses focus on characterizing bridge sets, or the set of bridge edges belonging to a given symptom. For example, $\mathcal{H}_1$ posits that the bridge set for node D8 is collectively greater than the set for node D6, with constraint that the edges within the bridge set for D8 are equal to each other. This effectively corresponds to testing whether node D8 has greater bridge strength than node D6. $\mathcal{H}_2$ then refines $\mathcal{H}_1$ by testing an exact order both between and within bridge sets.

\begin{align}
\label{eq:intra-inter}
    \mathcal{H}_1 &: \rho_{D8-A5} = \rho_{D8-A7} > (\rho_{D6-A3}, \rho_{D6-A6}) > 0 \\ \nonumber
    \mathcal{H}_2 &: \rho_{D8-A5} > \rho_{D8-A7} > \rho_{D6-A3} > \rho_{D6-A6} > 0 \\ \nonumber
    \mathcal{H}_3 &: ``\text{not}\; \mathcal{H}_1 \; \text{or}\; \mathcal{H}_2 \text{.''}
\end{align}

Note that the inclusion of an inequality and equality constraint in a single hypothesis is currently only available on the GitHub version for `BGGM`.

```{r}
intra_inter_hyp <- c("D8--A5 = D8--A7 > (D6--A3, D6--A6) > 0;
                      D8--A5 > D8--A7 > D6--A3 > D6--A6 > 0")

confirm_intra_inter <- confirm(Y = confirm_anxdep,
                               hypothesis = intra_inter_hyp,
                               iter = 50000)
confirm_intra_inter 
```

The data were more likely under both $\mathcal{H}_1$ ($\textbf{BF}_{13} = 4.5$) and $\mathcal{H}_2$ ($\textbf{BF}_{23} = 107$) than $\mathcal{H}_3$. Furthermore, there was more evidence supporting the hypothesis testing solely inequality constraints, $\mathcal{H}_2$, than the one including an equality constraint ($\textbf{BF}_{21} = 110.6$). This provides a clear characterization of the the bridge relations at hand --- not only did the order of bridge strength replicate, but so did the order of the edges within each bridge set. 

#### Bridge Set Separation

We also thought it would be interesting to test whether bridge sets include common elements. That is, whether bridge symptoms connect to the same or different nodes. As can be seen above, the bridge sets for nodes D8 and D6 are mutually exclusive. We can then test, say, whether D8 is conditionally independent the bridge set for D6 (nodes A3 and A6)

\begin{align}
    \mathcal{H}_1 &: (\rho_{D8-A3}, \rho_{D8-A6}) = 0 \\ \nonumber
    \mathcal{H}_2 &: (\rho_{D8-A3}, \rho_{D8-A6})  > 0 \\ \nonumber
    \mathcal{H}_3 &: ``\text{not}\; \mathcal{H}_1 \; \text{or} \; \mathcal{H}_2 \text{.''}
\end{align}


```{r include=F, eval=F}
hyp_bridge_sep <- ("(D8--A3, D8--A6) = 0;
                    (D8--A3, D8--A6) > 0")

confirm_bridge_sep <- confirm(Y = confirm_anxdep,
                      hypothesis = hyp_bridge_sep,
                      iter = 50000)
confirm_bridge_sep
```

Although the data were more likely under $\mathcal{H}_1$ than $\mathcal{H}_3$ (BF$_{13} = 3.7$), there was support in favor of $\mathcal{H}_2$ compared to $\mathcal{H}_1$ (BF$_{21} = 2$). This analysis suggests it is unlikely that D8 is actually conditionally independent from the bridge set for D6. 